Overt attentional capture by reward-related stimuli overcomes inhibitory suppression

### 2019-04-23 analysis/paper/paper.tex
fixed paper, added dockerfile for testing

\documentclass[jou, a4paper, noextraspace,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Overt attentional capture by reward-related stimuli overcomes inhibitory suppression},
            pdfauthor={Daniel Pearson, Poppy Watson, Phillip Cheng, \& Mike E. Le Pelley},
            pdfkeywords={attentional capture, reward, suppression, cognitive control},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

When we interact with the world, our visual attention system can
prioritise important information. For example, when driving towards an
intersection, we can selectively attend to a stop sign on a busy street.
Sometimes, however, we may find that stimuli that are irrelevant to our
current task may nevertheless capture our attention in a seemingly
automatic and involuntary way (e.g., a flashing alert, or colourful
billboard). Consistent with this framework, an influential view of
attentional selection has drawn a distinction between two forms of
attentional control: one that is volitional and goal directed (top-down
control) and another that is automatic and stimulus driven (S. Yantis,
2000). However, exactly how these two processes interact has been the
subject of extensive debate.

Most models of attentional selection can be categorised into one of two
opposing theoretical positions. \emph{Stimulus-driven} accounts argue
that attention is initially captured by the most salient stimulus
presented to an observer, regardless of their current goals (J.
Theeuwes, 1992, 2010). That is, individuals have no voluntary control
over the initial allocation of their attention, which is automatically
drawn to the most salient stimulus in a scene. By contrast,
\emph{goal-directed} accounts posit that attention is allocated only to
stimuli that are in some way related to the observer's current goals
(Folk, Remington, \& Johnston, 1992; Leber \& Egeth, 2006). These
theories predict that physically salient stimuli will be completely
ignored by the visual attention system, unless they share some
feature(s) with the observer's template of expected targets.

@@

For example, across two experiments Gaspelin et al. (2017) had
participants perform variants of a commonly used procedure for examining
attentional capture---the \emph{additional singleton task} (J. Theeuwes,
1991, 1992)---while recording eye movements. Eye movements provide a
useful index of attention since an eye movement to a location (a
saccade) is necessarily preceded by a shift of spatial attention to that
location (Deubel \& Schneider, 1996). In this task, participants were
required to locate and respond to a shape-singleton target that was
presented in an array of non-target shapes. On half of the trials, one
of the non-target shapes was rendered in a different colour from the
rest of the array so as to be a physically salient colour-singleton
distractor. If participants were \emph{more} likely to make a saccade
towards the singleton distractor than one of the nonsingleton
non-targets, this would indicate that the singleton distractor had
captured attention. By contrast, if participants were \emph{less} likely
to make a saccade to the singleton distractor than one of the
nonsingleton non-targets, this would suggest that attention to the
singleton distractor had been suppressed. In Gaspelin et al. (2017)'s
Experiment 1, participants were encouraged to adopt a strategy of
searching for the \enquote{odd-one-out} in the display (i.e,
\emph{singleton-search mode}; Bacon \& Egeth, 1994), by having the
target be randomly either a circle amongst diamonds or a diamond amongst
circles on each trial (see Figure \ref{fig:methodFig}A for an example
from the current study). Under these conditions, both stimulus-driven
(J. Theeuwes, 1992) and goal-directed (Bacon \& Egeth, 1994) accounts of
attentional selection would predict capture, as the template that
participants use to locate the target (i.e., any singleton) is broad
enough to include the singleton distractor. As expected, under these
singleton-search conditions Gaspelin et al. found that participants'
overt attention was captured by the salient singleton. In Experiment 2,
participants were instead encouraged to adopt a more specific target
template (i.e., \emph{feature-search mode}; Bacon \& Egeth, 1994), by
consistently mapping the target identity to one shape (e.g., search for
a circle on every trial), and having a heterogeneous set of shapes act
as the non-targets (see Figure \ref{fig:methodFig}B). This manipulation
makes searching for any singleton a highly inefficient strategy, as
there are multiple singleton nontargets in the display (e.g., a circle,
a square, a triangle, and a red shape). Thus, participants should be
encouraged to search for the specific features that define the target
(i.e., the \enquote{circleness} of the circle). Under these conditions,
attention to the salient colour-singleton distractor was suppressed
relative to the other non-targets, suggesting that the visual attention
system treated the salient singleton differently from the other
non-targets (inconsistent with goal-directed accounts), yet did not
capture gaze (inconsistent with stimulus-driven accounts). By contrast,
the signal suppression hypothesis can account for these findings by
suggesting that the singleton generated an attentional priority signal
as a consequence of its physical salience, but this signal could be
suppressed by an inhibitory mechanism under conditions that promoted a
more specific search strategy (i.e., feature search). Importantly, this
suppression effect was observed even for the most rapidly generated eye
movements, which was taken as evidence that the suppression effect
operates early in processing, before attention is initially allocated.

@@

Up to this point we have discussed the set of influences on attentional
control as a dichotomy: goal-directed versus stimulus-driven. However, a
growing literature has questioned this framework, pointing to
demonstrations of attentional selection biases that fall outside this
dichotomy (Awh, Belopolsky, \& Theeuwes, 2012; Failing \& Theeuwes,
2018; M. E. Le Pelley, Mitchell, Beesley, George, \& Wills, 2016). Awh
et al. (2012) suggested a third category of influences on attentional
selection, labelled \emph{selection history}. This category describes
attentional biases that are formed through learned relationships between
stimulus features and motivationally significant outcomes---such as
reward (Brian A. Anderson, 2016; Failing \& Theeuwes, 2018; M. E. Le
Pelley et al., 2016) or punishment (Watson, Pearson, Wiers, \& Le
Pelley, 2019)---as well as other biases related to the history of
attentional deployment that cannot be easily explained in terms of
goal-directed or stimulus-driven control, such as biases for selecting
stimulus features that have previously acted as targets (e.g., Maljkovic
\& Nakayama, 1994).

@@

Pairing a stimulus with reward is thought to increase its attentional
priority in a way that mimics or otherwise interacts with its physical
salience. For example, eye-tracking studies have demonstrated that VMAC
is largest for the most rapidly generated saccades, and diminishes with
increasing saccade latency (Failing, Nissens, Pearson, Le Pelley, \&
Theeuwes, 2015; Pearson et al., 2016), a pattern similar to that
observed in capture by physical salience (e.g., Zoest, Donk, \&
Theeuwes, 2004). Similarly, electrophysiological studies have shown that
reward affects ERP components thought to index low-level perceptual
activity (Hickey, Chelazzi, \& Theeuwes, 2010). This is consistent with
the idea of \emph{incentive salience} (Berridge \& Robinson, 1998),
wherein a reward-related stimulus becomes more salient and attention
grabbing to the observer through a change in its perceptual
representation. This raises the question, if associating a stimulus with
reward increases its perceptual priority as though it has become more
physically salient, can this priority signal be suppressed by the
inhibitory control process proposed by the signal suppression
hypothesis? That is, can incentive salience be suppressed in the same
way as physical salience? Aside from the theoretical importance of this
question, it may also have important clinical implications. Recent
evidence suggests that the processes underlying VMAC are closely related
to the attentional biases towards drug-related stimuli observed in
addiction (Albertella et al., 2017; Brian A Anderson, Faulkner, Rilee,
Yantis, \& Marvel, 2013; for review see Field \& Cox, 2008), which can
promote relapse in recovering addicts (Marhe, Waters, Wetering, \&
Franken, 2013; Marissen et al., 2006; A. J. Waters, Marhe, \& Franken,
2012). Thus it is important to know whether cognitive control processes
can be used to overcome the influence of reward on attention when such
capture is counterproductive and contrary to our goals. We addressed
this question by combining Le Pelley et al.'s VMAC procedure with
Gaspelin et al.'s (2017) overt attentional suppression procedure to
investigate whether reward-related attentional biases can be suppressed
by inhibitory control processes under conditions promoting feature
search.

@@

This study was approved by the UNSW Sydney Human Research Ethics
Advisory Panel (Psychology). Previous studies of VMAC using gaze
measures (M. E. Le Pelley et al., 2015; Pearson et al., 2015, 2016) have
demonstrated medium to large effect sizes (\(d_z\) = 0.41--1.4, mean =
0.73), and previous studies of overt attentional suppression of physical
salience (Gaspelin et al., 2017) have demonstrated a large effect size
(\(d_z\) = 1.6). Based on an anticipated effect size of \(d\) = 0.73, a
power analysis conducted with G*Power (Faul, Erdfelder, Lang, \&
Buchner, 2007) indicated that 32 participants per group would provide
adequate power (power \textasciitilde{}.98 to detect within-subjects
effects and power of \textasciitilde{}.82 for between-subjects effects).
We therefore tested 64 UNSW Sydney students (\(n =\) 32 per group, 44
females, age \(M=\) 20.15, \(SD=\) 4.74) who participated in exchange
for course credit or payment of 20 AUD. All participants also received a
monetary bonus that was dependent on their performance (\(M =\) 8.88
AUD, \(SD =\) 2.68 AUD).

@@

Participants were tested using a Tobii TX-300 eye-tracker (sampling
frequency 300 Hz), mounted on a 23-inch monitor (1920 \(\times\) 1080
resolution, 60 Hz refresh rate). Participants' heads were positioned in
a chin rest 60 cm from the screen. Gaze data were down-sampled to 100 Hz
for gaze contingent calculations during stimulus presentation, with gaze
position in each sample defined as the average position during the
preceding 10 ms window. Stimulus presentation was controlled by MATLAB
using the Psychophysics Toolbox extensions (D. H. Brainard, 1997;
Kleiner et al., 2007; Pelli, 1997). Experiment scripts, analysis
scripts, and raw data are available at (\textbf{OSF}).

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/Methods_fig}

@@

In line with previous VMAC protocols (e.g., M. E. Le Pelley et al.,
2015; Pearson et al., 2016), data from the first two experimental
trials, and the first two trials after each break were discarded. Hard
timeouts (2.5\% of all trials) were also discarded. One participant was
replaced because their mean proportion of valid gaze samples during each
trial was less than 50\%. One other participant was replaced after
reporting that they had misunderstood the task instructions. For the
remaining participants, valid gaze data were registered on an average of
97.5\% (\(SD=\) 3.7\%) of samples, suggesting high fidelity of gaze
data.

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/omission_plot}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/saccade_plot}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/capture_plot_by_block}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/vincentized_plot}

@@

In this experiment, we investigated the role of inhibitory control
processes in actively suppressing overt attention to physically salient
and reward-associated distractor stimuli. In line with the signal
suppression hypothesis (Gaspelin et al., 2015, 2017), overt attention to
physically salient colour singleton distractors was suppressed under
conditions that promoted top-down guidance of attention. Specifically,
under feature search conditions, participants were generally less likely
to look at a (low-reward) colour singleton distractor than the average
nonsingleton distractor. Consistent with recent findings (Gaspelin \&
Luck, 2018a; Vatterott \& Vecera, 2012), this attentional suppression
effect gradually emerged over the course of the experiment, as
participants gained experience with the salient-but-irrelevant
distractor feature (Figure \ref{fig:blockPlot}). Furthermore, we
observed an effect of reward association on overt attention (i.e., VMAC;
M. E. Le Pelley et al., 2015; Pearson et al., 2015), with participants'
gaze more likely to be captured by distractor stimuli that signalled
high reward than low reward, even though looking at reward-signalling
distractors caused omission of rewards that otherwise would have been
delivered. Similar to the attentional suppression effect, attentional
capture by reward-associated distractors also emerged over the course of
the experiment, as participants experienced more pairings of stimulus
features with reward. Critically, this study was the first to
demonstrate that---unlike the influence of physical salience---the
influence of reward association on attention cannot be suppressed under
conditions that promote top-down guidance of attention. Regardless of
whether participants were encouraged to search for specific target
defining features (i.e., feature search mode) or to search for the
odd-one-out (singleton search mode), stimuli associated with high reward
captured overt attention. Notably, this attentional bias was
counterproductive, in that it led to the omission of more high rewards
than low rewards, and so participants should have been highly motivated
to suppress their attention to the high-reward distractor if possible.
Nevertheless, the high-reward distractor captured attention, even under
conditions that allowed participants to suppress their attention to
physically salient distractors that were associated with low reward. Our
key finding, then, is that reward-related stimuli constitute especially
potent attentional priority signals that are relatively immune to
suppression by inhibitory control processes that can act to prevent
distraction by physical salience.

Most influential models of attentional selection assume that
stimulus-driven and goal-directed influences are integrated onto a
common priority map in the brain, with attention directed towards the
spatial location with the highest peak of activation (e.g., Itti \&
Koch, 2001; J. Theeuwes, 2010; Wolfe, 2007). The priority map concept
has also been used to explain how selection history influences
attentional allocation (Awh et al., 2012; Failing \& Theeuwes, 2018),
with the attentional priority signal of reward-related stimulus features
increasing with repeated stimulus-reward pairings through a process of
associative learning (M. E. Le Pelley et al., 2016; Pearson et al.,
2016). Recent formulations of the signal suppression hypothesis have
suggested that the attentional suppression mechanism may operate in a
similar way, with the attentional priority of to-be-ignored stimulus
features being proactively down-weighted with repeated experience
(Gaspelin \& Luck, 2019). At first glance, the results of the current
study may seem inconsistent with the idea that reward and suppression
engage in competition on a common priority map, as the magnitude of the
VMAC effect did not differ between search conditions, even though the
feature search condition should have encouraged top-down guidance of
attention. However, there was an \emph{overall} decrease in the amount
of capture by both the high- and low-reward distractor in the feature
search condition (supported by a main effect of search condition, e.g.,
see Figures \ref{fig:SaccadePlot}B and \ref{fig:SaccadePlot}D). This
raises the possibility that repeated experience of the physically
salient stimulus features (i.e., orange and blue colour) allows the
attentional priority of those features to be uniformly down-weighted by
the suppression mechanism. In addition, our findings are consistent with
the idea that repeated experience of pairings between stimulus features
and reward results in selective up-weighting of the attentional priority
of the high-reward distractor, such that it overpowers the influence of
inhibitory suppression and thus captures attention.

>>>>

\documentclass[man, a4paper, noextraspace, 11pt,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Overt attentional capture by reward-related stimuli overcomes inhibitory suppression},
            pdfauthor={Daniel Pearson, Poppy Watson, Phillip Cheng, \& Mike E. Le Pelley},
            pdfkeywords={attentional capture, reward, suppression, cognitive control},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

When we interact with the world, our visual attention system can
prioritise important information. For example, when driving towards an
intersection, we can selectively attend to a stop sign on a busy street.
Sometimes, however, we may find that stimuli that are irrelevant to our
current task may nevertheless capture our attention in a seemingly
automatic and involuntary way (e.g., a flashing alert, or colourful
billboard). Consistent with this framework, an influential view of
attentional selection has drawn a distinction between two forms of
attentional control: one that is volitional and goal directed (top-down
control) and another that is automatic and stimulus driven (Yantis,
2000). However, exactly how these two processes interact has been the
subject of extensive debate.

Most models of attentional selection can be categorised into one of two
opposing theoretical positions. \emph{Stimulus-driven} accounts argue
that attention is initially captured by the most salient stimulus
presented to an observer, regardless of their current goals (Theeuwes,
1992, 2010). That is, individuals have no voluntary control over the
initial allocation of their attention, which is automatically drawn to
the most salient stimulus in a scene. By contrast, \emph{goal-directed}
accounts posit that attention is allocated only to stimuli that are in
some way related to the observer's current goals (Folk, Remington, \&
Johnston, 1992; Leber \& Egeth, 2006). These theories predict that
physically salient stimuli will be completely ignored by the visual
attention system, unless they share some feature(s) with the observer's
template of expected targets.

@@

For example, across two experiments Gaspelin et al. (2017) had
participants perform variants of a commonly used procedure for examining
attentional capture---the \emph{additional singleton task} (Theeuwes,
1991, 1992)---while recording eye movements. Eye movements provide a
useful index of attention since an eye movement to a location (a
saccade) is necessarily preceded by a shift of spatial attention to that
location (Deubel \& Schneider, 1996). In this task, participants were
required to locate and respond to a shape-singleton target that was
presented in an array of non-target shapes. On half of the trials, one
of the non-target shapes was rendered in a different colour from the
rest of the array so as to be a physically salient colour-singleton
distractor. If participants were \emph{more} likely to make a saccade
towards the singleton distractor than one of the nonsingleton
non-targets, this would indicate that the singleton distractor had
captured attention. By contrast, if participants were \emph{less} likely
to make a saccade to the singleton distractor than one of the
nonsingleton non-targets, this would suggest that attention to the
singleton distractor had been suppressed. In Gaspelin et al. (2017)'s
Experiment 1, participants were encouraged to adopt a strategy of
searching for the \enquote{odd-one-out} in the display (i.e,
\emph{singleton-search mode}; Bacon \& Egeth, 1994), by having the
target be randomly either a circle amongst diamonds or a diamond amongst
circles on each trial (see Figure \ref{fig:methodFig}A for an example
from the current study). Under these conditions, both stimulus-driven
(Theeuwes, 1992) and goal-directed (Bacon \& Egeth, 1994) accounts of
attentional selection would predict capture, as the template that
participants use to locate the target (i.e., any singleton) is broad
enough to include the singleton distractor. As expected, under these
singleton-search conditions Gaspelin et al. found that participants'
overt attention was captured by the salient singleton. In Experiment 2,
participants were instead encouraged to adopt a more specific target
template (i.e., \emph{feature-search mode}; Bacon \& Egeth, 1994), by
consistently mapping the target identity to one shape (e.g., search for
a circle on every trial), and having a heterogeneous set of shapes act
as the non-targets (see Figure \ref{fig:methodFig}B). This manipulation
makes searching for any singleton a highly inefficient strategy, as
there are multiple singleton nontargets in the display (e.g., a circle,
a square, a triangle, and a red shape). Thus, participants should be
encouraged to search for the specific features that define the target
(i.e., the \enquote{circleness} of the circle). Under these conditions,
attention to the salient colour-singleton distractor was suppressed
relative to the other non-targets, suggesting that the visual attention
system treated the salient singleton differently from the other
non-targets (inconsistent with goal-directed accounts), yet did not
capture gaze (inconsistent with stimulus-driven accounts). By contrast,
the signal suppression hypothesis can account for these findings by
suggesting that the singleton generated an attentional priority signal
as a consequence of its physical salience, but this signal could be
suppressed by an inhibitory mechanism under conditions that promoted a
more specific search strategy (i.e., feature search). Importantly, this
suppression effect was observed even for the most rapidly generated eye
movements, which was taken as evidence that the suppression effect
operates early in processing, before attention is initially allocated.

@@

Up to this point we have discussed the set of influences on attentional
control as a dichotomy: goal-directed versus stimulus-driven. However, a
growing literature has questioned this framework, pointing to
demonstrations of attentional selection biases that fall outside this
dichotomy (Awh, Belopolsky, \& Theeuwes, 2012; Failing \& Theeuwes,
2018; Le Pelley, Mitchell, Beesley, George, \& Wills, 2016). Awh et al.
(2012) suggested a third category of influences on attentional
selection, labelled \emph{selection history}. This category describes
attentional biases that are formed through learned relationships between
stimulus features and motivationally significant outcomes---such as
reward (Anderson, 2016; Failing \& Theeuwes, 2018; Le Pelley et al.,
2016) or punishment (Watson, Pearson, Wiers, \& Le Pelley, 2019)---as
well as other biases related to the history of attentional deployment
that cannot be easily explained in terms of goal-directed or
stimulus-driven control, such as biases for selecting stimulus features
that have previously acted as targets (e.g., Maljkovic \& Nakayama,
1994).

@@

Pairing a stimulus with reward is thought to increase its attentional
priority in a way that mimics or otherwise interacts with its physical
salience. For example, eye-tracking studies have demonstrated that VMAC
is largest for the most rapidly generated saccades, and diminishes with
increasing saccade latency (Failing, Nissens, Pearson, Le Pelley, \&
Theeuwes, 2015; Pearson et al., 2016), a pattern similar to that
observed in capture by physical salience (e.g., Zoest, Donk, \&
Theeuwes, 2004). Similarly, electrophysiological studies have shown that
reward affects ERP components thought to index low-level perceptual
activity (Hickey, Chelazzi, \& Theeuwes, 2010). This is consistent with
the idea of \emph{incentive salience} (Berridge \& Robinson, 1998),
wherein a reward-related stimulus becomes more salient and attention
grabbing to the observer through a change in its perceptual
representation. This raises the question, if associating a stimulus with
reward increases its perceptual priority as though it has become more
physically salient, can this priority signal be suppressed by the
inhibitory control process proposed by the signal suppression
hypothesis? That is, can incentive salience be suppressed in the same
way as physical salience? Aside from the theoretical importance of this
question, it may also have important clinical implications. Recent
evidence suggests that the processes underlying VMAC are closely related
to the attentional biases towards drug-related stimuli observed in
addiction (Albertella et al., 2017; Anderson, Faulkner, Rilee, Yantis,
\& Marvel, 2013; for review see Field \& Cox, 2008), which can promote
relapse in recovering addicts (Marhe, Waters, Wetering, \& Franken,
2013; Marissen et al., 2006; Waters, Marhe, \& Franken, 2012). Thus it
is important to know whether cognitive control processes can be used to
overcome the influence of reward on attention when such capture is
counterproductive and contrary to our goals. We addressed this question
by combining Le Pelley et al.'s VMAC procedure with Gaspelin et al.'s
(2017) overt attentional suppression procedure to investigate whether
reward-related attentional biases can be suppressed by inhibitory
control processes under conditions promoting feature search.

@@

This study was approved by the UNSW Sydney Human Research Ethics
Advisory Panel (Psychology). Previous studies of VMAC using gaze
measures (Le Pelley et al., 2015; Pearson et al., 2015, 2016) have
demonstrated medium to large effect sizes (\(d_z\) = 0.41--1.4, mean =
0.73), and previous studies of overt attentional suppression of physical
salience (Gaspelin et al., 2017) have demonstrated a large effect size
(\(d_z\) = 1.6). Based on an anticipated effect size of \(d\) = 0.73, a
power analysis conducted with G*Power (Faul, Erdfelder, Lang, \&
Buchner, 2007) indicated that 32 participants per group would provide
adequate power (power \textasciitilde{}.98 to detect within-subjects
effects and power of \textasciitilde{}.82 for between-subjects effects).
We therefore tested 64 UNSW Sydney students (\(n =\) 32 per group, 44
females, age \(M=\) 20.15, \(SD=\) 4.74) who participated in exchange
for course credit or payment of 20 AUD. All participants also received a
monetary bonus that was dependent on their performance (\(M =\) 8.88
AUD, \(SD =\) 2.68 AUD).

@@

Participants were tested using a Tobii TX-300 eye-tracker (sampling
frequency 300 Hz), mounted on a 23-inch monitor (1920 \(\times\) 1080
resolution, 60 Hz refresh rate). Participants' heads were positioned in
a chin rest 60 cm from the screen. Gaze data were down-sampled to 100 Hz
for gaze contingent calculations during stimulus presentation, with gaze
position in each sample defined as the average position during the
preceding 10 ms window. Stimulus presentation was controlled by MATLAB
using the Psychophysics Toolbox extensions (Brainard, 1997; Kleiner et
al., 2007; Pelli, 1997). Experiment scripts, analysis scripts, and raw
data are available at (\textbf{OSF}).

@@

\begin{figure}[!h]

{\centering \includegraphics[width=0.8\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/Methods_fig}

@@

In line with previous VMAC protocols (e.g., Le Pelley et al., 2015;
Pearson et al., 2016), data from the first two experimental trials, and
the first two trials after each break were discarded. Hard timeouts
(2.5\% of all trials) were also discarded. One participant was replaced
because their mean proportion of valid gaze samples during each trial
was less than 50\%. One other participant was replaced after reporting
that they had misunderstood the task instructions. For the remaining
participants, valid gaze data were registered on an average of 97.5\%
(\(SD=\) 3.7\%) of samples, suggesting high fidelity of gaze data.

@@

\begin{figure}[!h]

{\centering \includegraphics[width=0.6\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/omission_plot}

@@

\begin{figure}[!h]

{\centering \includegraphics[width=0.8\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/saccade_plot}

@@

\begin{figure}[!h]

{\centering \includegraphics[width=0.6\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/capture_plot_by_block}

@@

\begin{figure}[!h]

{\centering \includegraphics[width=0.6\linewidth]{C:/Users/Daniel/Google Drive/Research/vmacfeaturesearch/analysis/figures/vincentized_plot}

@@

In this experiment, we investigated the role of inhibitory control
processes in actively suppressing overt attention to physically salient
and reward-associated distractor stimuli. In line with the signal
suppression hypothesis (Gaspelin et al., 2015, 2017), overt attention to
physically salient colour singleton distractors was suppressed under
conditions that promoted top-down guidance of attention. Specifically,
under feature search conditions, participants were generally less likely
to look at a (low-reward) colour singleton distractor than the average
nonsingleton distractor. Consistent with recent findings (Gaspelin \&
Luck, 2018a; Vatterott \& Vecera, 2012), this attentional suppression
effect gradually emerged over the course of the experiment, as
participants gained experience with the salient-but-irrelevant
distractor feature (Figure \ref{fig:blockPlot}). Furthermore, we
observed an effect of reward association on overt attention (i.e., VMAC;
Le Pelley et al., 2015; Pearson et al., 2015), with participants' gaze
more likely to be captured by distractor stimuli that signalled high
reward than low reward, even though looking at reward-signalling
distractors caused omission of rewards that otherwise would have been
delivered. Similar to the attentional suppression effect, attentional
capture by reward-associated distractors also emerged over the course of
the experiment, as participants experienced more pairings of stimulus
features with reward. Critically, this study was the first to
demonstrate that---unlike the influence of physical salience---the
influence of reward association on attention cannot be suppressed under
conditions that promote top-down guidance of attention. Regardless of
whether participants were encouraged to search for specific target
defining features (i.e., feature search mode) or to search for the
odd-one-out (singleton search mode), stimuli associated with high reward
captured overt attention. Notably, this attentional bias was
counterproductive, in that it led to the omission of more high rewards
than low rewards, and so participants should have been highly motivated
to suppress their attention to the high-reward distractor if possible.
Nevertheless, the high-reward distractor captured attention, even under
conditions that allowed participants to suppress their attention to
physically salient distractors that were associated with low reward. Our
key finding, then, is that reward-related stimuli constitute especially
potent attentional priority signals that are relatively immune to
suppression by inhibitory control processes that can act to prevent
distraction by physical salience.

Most influential models of attentional selection assume that
stimulus-driven and goal-directed influences are integrated onto a
common priority map in the brain, with attention directed towards the
spatial location with the highest peak of activation (e.g., Itti \&
Koch, 2001; Theeuwes, 2010; Wolfe, 2007). The priority map concept has
also been used to explain how selection history influences attentional
allocation (Awh et al., 2012; Failing \& Theeuwes, 2018), with the
attentional priority signal of reward-related stimulus features
increasing with repeated stimulus-reward pairings through a process of
associative learning (Le Pelley et al., 2016; Pearson et al., 2016).
Recent formulations of the signal suppression hypothesis have suggested
that the attentional suppression mechanism may operate in a similar way,
with the attentional priority of to-be-ignored stimulus features being
proactively down-weighted with repeated experience (Gaspelin \& Luck,
2019). At first glance, the results of the current study may seem
inconsistent with the idea that reward and suppression engage in
competition on a common priority map, as the magnitude of the VMAC
effect did not differ between search conditions, even though the feature
search condition should have encouraged top-down guidance of attention.
However, there was an \emph{overall} decrease in the amount of capture
by both the high- and low-reward distractor in the feature search
condition (supported by a main effect of search condition, e.g., see
Figures \ref{fig:SaccadePlot}B and \ref{fig:SaccadePlot}D). This raises
the possibility that repeated experience of the physically salient
stimulus features (i.e., orange and blue colour) allows the attentional
priority of those features to be uniformly down-weighted by the
suppression mechanism. In addition, our findings are consistent with the
idea that repeated experience of pairings between stimulus features and
reward results in selective up-weighting of the attentional priority of
the high-reward distractor, such that it overpowers the influence of
inhibitory suppression and thus captures attention.

An inconsistency between the results of the current study and those of
Gaspelin et al. (2017) comes from the time-course of the attentional
capture and suppression effects that were observed. According to the
signal suppression hypothesis, suppression is \emph{proactive}---in that
it operates prior to the initial allocation of attention---and so
suppression should be observed even for the most rapidly generated
saccades. Consistent with this idea, Gaspelin et al. (2017) observed an
overt attentional suppression effect to physically salient singleton
distractors amongst the fastest quartile of saccade latencies. By
contrast, in the current experiment we observed overt attentional
capture in the fastest quartile of saccade latencies, with the
suppression effect emerging on trials with slower saccades.
Methodological differences between Gaspelin et al.\enquote{s (2017)
experiment and the current study may be responsible for this
discrepancy. For example, the availability of reward on low-reward
trials may have influenced capture, either by somewhat increasing the
attentional priority of the stimulus feature associated with low reward,
or by otherwise influencing participants} motivation to respond quickly.
Alternatively, the discrepancy in findings may reflect our use of two
potential targets (circle and diamond) in the feature search condition,
as opposed to the same target on all trials in Gaseplin et al.'s
procedure. Having two possible targets could reduce the influence of
trial-by-trial upweighting of target features (Maljkovic \& Nakayama,
1994), but may also result in less effective feature search (but see
Grubert \& Eimer, 2015; e.g., Houtkamp \& Roelfsema, 2009). Another
possible explanation comes from the use of an intermixed trial
structure: as participants randomly experienced trials containing either
the high- or low-reward colour throughout the course of the task, it is
possible that the proactive down-weighting of the salient-but-irrelevant
features could not build up enough to prevent capture for the most
rapidly initiated saccades (Gaspelin \& Luck, 2019). How these
methodological differences might influence the time-course of overt
attentional capture and suppression remains a question for future
research. Nonetheless, the finding does raise questions about whether
the attentional suppression effect that we have observed is truly the
result of proactive attentional suppression of salient stimuli, as
suggested by the signal suppression hypothesis, or potentially the
result of rapid covert disengagement from the salient stimuli (see
Theeuwes, 2010; Zoest et al., 2004 for similar arguments).
